Plasticity and adaptation to high light intensity amplify the advantage of amphistomatous leaves

### 2025-08-14 ms/ms.tex
proofread supporting materials

\bibitem[\citeproctext]{ref-schoch_dependence_1980}
\CSLLeftMargin{54. }%
\CSLRightInline{P.-G. Schoch, C. Zinsou, M. Sibi,
\href{https://doi.org/10.1093/jxb/31.5.1211}{Dependence of the stomatal
index on environmental factors during stomatal differentiation in leaves
of \emph{{Vigna} sinensis} {L}.: 1. {Effect} of light intensity}.
\emph{Journal of Experimental Botany} \textbf{31}, 1211--1216 (1980).}

\bibitem[\citeproctext]{ref-sack_developmental_2016}
\CSLLeftMargin{55. }%
\CSLRightInline{L. Sack, T. N. Buckley,
\href{https://doi.org/10.1104/pp.16.00476}{The developmental basis of
stomatal density and flux}. \emph{Plant Physiology} \textbf{171},
2358--2363 (2016).}

\bibitem[\citeproctext]{ref-mott_amphistomy_1991}
\CSLLeftMargin{56. }%
\CSLRightInline{K. A. Mott, O. Michaelson, Amphistomy as an adaptation
to high light intensity in \emph{{Ambrosia} cordifolia} ({Compositae}).
\emph{American Journal of Botany} \textbf{78}, 76--79 (1991).}

\bibitem[\citeproctext]{ref-schindelin_fiji_2012}
\CSLLeftMargin{57. }%
\CSLRightInline{J. Schindelin, I. Arganda-Carreras, E. Frise, V. Kaynig,
M. Longair, T. Pietzsch, S. Preibisch, C. Rueden, S. Saalfeld, B.
Schmid, J.-Y. Tinevez, D. J. White, V. Hartenstein, K. Eliceiri, P.
Tomancak, A. Cardona, \href{https://doi.org/10.1038/nmeth.2019}{Fiji: An
open-source platform for biological-image analysis}. \emph{Nature
Methods} \textbf{9}, 676--682 (2012).}

\bibitem[\citeproctext]{ref-monteith_principles_2013}
\CSLLeftMargin{58. }%
\CSLRightInline{J. L. Monteith, M. H. Unsworth, \emph{Principles of
Environmental Physics: Plants, Animals, and the Atmosphere}
(Elsevier/Academic Press, Amsterdam ; Boston, 4th ed., 2013).}

\bibitem[\citeproctext]{ref-stan_development_team_stan_2025}
\CSLLeftMargin{59. }%
\CSLRightInline{Stan Development Team, \emph{Stan {Modeling} {Language}
{Users} {Guide} and {Reference} {Manual}} (2025;
\url{https://mc-stan.org}).}

\bibitem[\citeproctext]{ref-burkner_brms_2017}
\CSLLeftMargin{60. }%
\CSLRightInline{P.-C. Bürkner,
\href{https://doi.org/10.18637/jss.v080.i01}{\textbf{Brms} : {An}
\emph{r} {Package} for {Bayesian} {Multilevel} {Models} {Using}
\emph{stan}}. \emph{Journal of Statistical Software} \textbf{80}
(2017).}

\bibitem[\citeproctext]{ref-gabry_cmdstanr_2025}
\CSLLeftMargin{61. }%
\CSLRightInline{J. Gabry, R. Češnovar, A. Johnson, S. Bronder,
\emph{Cmdstanr: {R} {Interface} to '{CmdStan}'} (2025;
\href{https://mc-stan.org/cmdstanr,\%20https://discourse.mc-stan.org}{https://mc-stan.org/cmdstanr,
https://discourse.mc-stan.org}).}

\bibitem[\citeproctext]{ref-r_core_team_r:_2025}
\CSLLeftMargin{62. }%
\CSLRightInline{R Core Team, \emph{R: {A} {Language} and {Environment}
for {Statistical} {Computing}} (R Foundation for Statistical Computing,
Vienna, Austria, 2025; \url{http://www.R-project.org/}).}

\bibitem[\citeproctext]{ref-gelman_inference_1992}
\CSLLeftMargin{63. }%
\CSLRightInline{A. Gelman, D. B. Rubin, Inference from iterative
simulation using multiple sequences. \emph{Statistical Science}
\textbf{7}, 457--472 (1992).}

\bibitem[\citeproctext]{ref-pease_phylogenomics_2016}
\CSLLeftMargin{64. }%
\CSLRightInline{J. B. Pease, D. C. Haak, M. W. Hahn, L. C. Moyle,
\href{https://doi.org/10.1371/journal.pbio.1002379}{Phylogenomics
reveals three sources of adaptive variation during a rapid radiation}.
\emph{PLOS Biology} \textbf{14}, e1002379 (2016).}

@@

We compared \aax{} among 29 ecologically diverse populations of wild
tomato, including representatives of all described species of
\emph{Solanum} sect. \emph{Lycopersicon} and sect.
\emph{Lycopersicoides} (\emph{50}) and the cultivated tomato \emph{S.
lycopersicum} var. \emph{lycopersicum} (Table~\ref{tbl-populations}).
Due to constraints on growth space and time, we spread out measurements
over 61.1 weeks from August 29, 2022 to October 31, 2023. Replicates
within population were evenly spread out over this period to prevent
confounding of temporal variation in growth conditions with accession.
{[}anything else to say here? maybe explain population selection and
phylogeny?{]}

@@

We recorded \(\mathrm{PPFD}\) using full spectrum quantum sensors
(SQ-500-SS, Apogee Instruments, Logan, Utah, USA); we recorded
temperature, RH, and {[}CO\(_2\){]} using an EE894 sensor (E+E
Elektronik, Engerwitzdorf, Austria) protected by a radiation shield. All
environmental measurements were taken every 10 minutes from the middle
of plants racks at approximately the same height as the leaves we
measured. We measured leaf temperature of focal leaves prior to
measurement using an infrared radiometer (SI-111-SS, Apogee Instruments,
Logan, Utah, USA).

@@

Seeds provided by the Tomato Genetics Resource Center germinated on
moist paper in plastic boxes after soaking for 30-60 minutes in a 50\%
(volume per volume) solution of household bleach and water, followed by
a thorough rinse. We transferred seedlings to cell-pack flats containing
Pro-Mix BX potting mix (Premier Tech, Rivière-du-Loup, Quebec, Canada)
once cotyledons fully emerged, typically within 1-2 weeks of sowing. We
grew seeds and seedlings for both sun and shade treatments under the
same environmental conditions (12:12 h,
24.3:\(\qty{21.7}{\degreeCelsius}\), 49.6:58.4 RH day:night cycle). LED
light provided
\(\mathrm{PPFD} = \qty{267}{\micro\mol\raiseto{-2}\meter\raiseto{-1}\second}\)
(Fluence RAZRx, Austin, Texas, USA).

@@

We selected a fully expanded, unshaded leaf at least six leaves above
the cotyledons during early vegetative growth. This typically meant that
plants had grown in light treatments for \(\approx\) 4 weeks, ensuring
they had time to sense and respond developmentally to the light
intensity of the treatment rather than the seedling conditions
(\emph{54}). Shade plants grew slower than sun plants, hence leaves at
the same developmental stage were measured on chronologically older
plants in the shade treatment. In some sun plants, we had to use leaves
higher on the stem because short internodes made lower leaves
inaccessible with the gas exchange equipment. We measured terminal
leaflets in 82.6\% of cases, but used the lateral leaflet closest to the
terminal leaflet when it was damaged or difficult to clamp into the gas
exchange chamber. When a leaflet was damaged during gas exchange
measurements, we collected anatomical data from the nearest leaflet on
the same leaf (1.58\% of leaves).

@@

\paragraph{Light-saturated photosynthetic
rate}\label{light-saturated-photosynthetic-rate}

In 91.3\% of plants, we measured the maximum photosynthetic rate
(\(A_\mathrm{max}\)) on the same leaflets as \agcurve{} curves. However,
when a leaflet was damaged during \agcurve{} curves, we used the next
closest leaflet for \(A_\mathrm{max}\). Leaves acclimated to high
light-intensity (\ppfdequals{2000}), ambient CO\(_2\) (\caequals{415}),
\rhequals{50}, and \tleafequals{25}. After \(A\) and \gsw{} stabilized,
we measured \(A\) at
\(\qty{2000}{\micro\mol\raiseto{-2}\meter\raiseto{-1}\second}\).

@@

We estimated the stomatal density and size on ab- and adaxial leaf
surfaces from all leaves, using guard cell length as a proxy for
stomatal size since it is proportional to maximum conductance
(\emph{55}). We made surface impressions of leaf lamina from the same
area used for gas exchange measurements using a-silicone impression
material (Zhermack elite HD+, light body, fast set, Rovigo, Italy). We
applied clear nail polish to make positive replicas of the impression.
After nail polish dried, we mounted replicas on a microscope slide using
transparent tape (\emph{56}). We digitized a portion of each leaf
surface replica using a brightfield microscope (Leica DM2000, Wetzlar,
Germany). We counted and measured guard cell length on all stomata using
the FIJI implementation of ImageJ2 version 2.3.0 (\emph{57}), then
divided the count by the visible leaf area
(\(\qty{0.890}{\raiseto{2}\milli\meter}\)) to estimate stomatal density.
For each surface we calculated the anatomical maximum stomatal
conductance (\gmax) at a reference leaf temperature of
\(\qty{25}{\degreeCelsius}\) following (\emph{55}) as:

@@

where \(D_\mathrm{wv}\) is the diffusion coefficient of water vapor in
air and \(v\) is by the kinematic viscosity of dry air. We assumed
\(D_\mathrm{wv} = \qty{2.49e-5}{\meter\squared\per\second}\) and
\(v = \qty{2.24e-2}{\raiseto{3}\meter\per\mol}\) (\emph{58}). For
kidney-shaped guard cells like those in wild tomatoes,
\(c = h = j = 0.5\). The \gmaxratio{} is the \gmax{} of the adaxial
surface divided by the sum total \gmax{} of both surfaces.

@@

Leaf mass per area (LMA) is the dry mass divided by the leaflet area. We
scanned fresh leaflets on a flat bed scanner (Epson V600, Los Alamitos,
California. USA) and measured leaflet area from digital images using the
FIJI implementation of ImageJ2 version 2.3.0 (\emph{57}). We dried
leaves for 72 hours at \(\qty{74}{\degreeCelsius}\) in a food dehydrator
(Cosori CP267-FD, Vesync Co., Anaheim, California, USA) and weighed
using a benchtop analytical balance (Ohaus PR64 Analytical Balance,
Parsippany, New Jersey, USA). In \(10.5\%\) of leaves we measured LMA on
the adjacent leaflet because the focal leaflet was damaged or wilted
while making surface impressions and we could not reliably estimate
area. LMA data are missing from \(2.97\%\) of individuals because the
area or mass was not recorded at all or recorded incorrectly.

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Estimate \agcurve{} curve parameters for each leaf
\item
  Estimate \aax{} for both measurement light intensities using
  \agcurve{} curve parameters
\item
  Estimate the effects of light intensity, light treatment, and
  population on \aax{} (acclimation and plasticity hypotheses)
\item
  Estimate the effects of native light habitat on population-level
  \aax{} (constitutive hypothesis)
\end{enumerate}

All models were using a Bayesian model with HMC sampling in the
probabilistic programming language \emph{Stan} (\emph{59}) using the
\emph{R} package \textbf{brms} version 2.22.0 (\emph{60}). We used
\emph{CmdStan} version 2.36.0 and \textbf{cmdstanr} version 0.9.0
(\emph{61}) to interface with \emph{R} version 4.5.1 (\emph{62}). We
sampled the posterior distribution from a single chain with a minimum of
1000 iterations after 1000 warmup iterations. If necessary, we refit
models with more iterations to decrease the Gelman-Rubin convergence
statistic (\(\hat{R}\)) (\emph{63}) to less than 1.01 and the bulk
effective sample size (ESS) to greater than 400 for all model
parameters, and there were fewer than 10 divergent transitions.

@@

We fit Bayesian mixed effects models with phylogenetically structured
random effects to test predictions of competing hypotheses about why
amphistomy advantage (\aax) might be greater for leaves in sunny, open
habitats. We fit all models combinations of measurement light intensity,
growth light intensity, and their interaction as both fixed effects and
random effects among populations. In other words, we test whether there
were main effects of light intensity treatments and/or whether
populations varied in their response to light intensity treatments. In
all models, we included population as a phylogenetically structured
random effect. We accounted for uncertainty in \(\widehat{\mathrm{AA}}\)
using the standard deviation of the posterior distribution of each
\(\widehat{\mathrm{AA}}\) estimate (Section~\ref{sec-aa-est}). We
crossed fixed and random effect structures with fixed effects of
measurement light intensity and/or growth light intensity on the
distributional parameter, \(\sigma\), the residual variance in \aax{}.
We used a robust regression approach by assuming \(t\)-distributed
residuals, which makes estimates less susceptible to extreme values. In
total, we fit 100 models, each with a different combination of fixed,
random, and distributional effects. We used the RAxML
whole-transcriptome concatenated phylogeny based on 2,745 100-kb genomic
windows from (\emph{64}). Two of our populations were not in this tree.
We used accession LA1044 in place of LA3909, two populations of \emph{S.
galapagnse}; LA0750 was added as sister to LA0716, two closely related
populations of \emph{S. pennellii}. The node separating LA0750 from
LA0716 was placed half-way between the next deepest node. We used the
leave-one-out cross-validation information criterion (LOOIC) to compare
the fit of models (\autoref{tbl-aa_loo1}) using the \emph{R} package
\textbf{loo} version 2.8.0 (\emph{65}) to calculate LOOIC values. We
selected the model with the lowest LOOIC value (Table~\ref{tbl-aa_loo1})
to generate posterior predictions for hypothesis testing. Parameter
estimates from the selected model are shown in Table~\ref{tbl-fit_aa1}.

To test whether LMA mediated the relationship between growth light
intensity and \aax{}, we repeated a similar model comparison procedure
but included \(\log \mathrm{LMA}\) as an additional fixed effect in the
model, as well its interaction with measurement and growth light
intensity. The model included a direct effect of growth light intensity
on \(\log \mathrm{LMA}\) and random phylogeneticlly structured effects
of population on \(\log \mathrm{LMA}\) and effects of sun treatment on
\(\log \mathrm{LMA}\). This enabled us to disentangle whether
\(\log \mathrm{LMA}\) effects \aax{} directly or indirectly because both
traits respond to growth light intensity. We integrated over missing LMA
data using the \texttt{mi()} function in \textbf{brms}. Because the
evidence strong supported including effects of growth and measurement
light intensity on the residual variance in \(\mathrm{AA}\), we included
these in terms in all models. In total, we fit 75 models, each with a
different combination of fixed and random effect structures. We selected
the model with the lowest LOOIC value (Table~\ref{tbl-aa_loo2}) to
generate posterior predictions for hypothesis testing. Note that LOOIC
was calculated only from pointwise likelihood values of the submodel
estimating effects of light intensity, light treatments, and population
on \aax{}. Parameter estimates from the selected model are shown in
Table~\ref{tbl-fit_aa2}.

@@

The acclimation, plasticity, and constitutive hypotheses make different
predictions about the relationship between \aax{} and light intensity,
light treatment, and native \ppfd{} among populations. Since these
hypotheses are not mutually exclusive, we describe how we assessed
support for one, two, or all three hypotheses simultaneously in
\autoref{tbl-predictions}. In general, the acclimation hypothesis was
supported if \aax{} was greater at high light intensity than low light
intensity. The plastic hypothesis was supported if \aax{} was greater in
sun leaves than shade leaves. The constitutive hypothesis was supported
if population-level \aax{} increased with native \ppfd. In interactive
models, we only consider positively reinforcing interactions between
high light intensity, sun leaves, and native \ppfd{} because these are
the only interactions which could explain why amphistomatous leaves are
advantageous in high light habitats.

From the posterior of the selected model (Section~\ref{sec-phylo}), we
predicted the posterior distribution of the expected \aax{} for each
population at each measurement light intensity and growth light
intensity. If the 95\% confidence intervals for \aax{} did not overlap
zero, we considered \aax{} to be significantly different from zero. We
also calculated posterior distribution for the average \aax{} among
populations in each condition. We tested a linear effect of native light
habitat on population-level \aax{} in each treatment condition,
integrating over the entire posterior distribution. We calculated the
posterior distribution of the slope. If the 95\% confidence interval for
the slope did not overlap zero, we considered the slope to be
significant.

@@

\subsubsection{Estimating the water cost of
hypostomy}\label{estimating-the-water-cost-of-hypostomy}

Amphistomy advantage implies that hypostomatous leaves would need to
lose more water by having higher \gsw{} to achieve the same
photosynthetic rate. We estimated this cost of hypostomy approximately.
For simpliticity, we focus on a sufficiently small region of the
\agcurve{} curve such that we can approximate the relationship between
\gsw{} and \(A\) as log-log linear. We also assume that slope is same in
both amphistomatous and hypostomatous leaves, which was often
approximately the case in our experiment (Figure~\ref{fig-curves}). With
these assumptions, the approximate water cost of hypostomy can be
derived from the equations:

@@

\[-\log\left(\frac{g_\text{sw,amphi}}{g_\text{sw,hypo}}\right) = \frac{\mathrm{AA}}{\varepsilon_g}.\]
We note that the slope in this case is an elasticity of of \(A\) to
\gsw{} because both are log-transformed. This quantity will be
proportional the water cost of hypostomy assuming that VPD and other
environmental variables are held constant and boundary layer conductance
are high enough to be negligible.

\subsection{Results}\label{results}

@@

\caption{\label{tbl-fit_aa2}Parameter estimates and 95\% confidence
intervals (CIs) from the posterior distribution of the selected model
with leaf mass per area (LMA) as potential mediator variable. Models
potentially include fixed, random, and distributional effects of growth
light intensity (shade and sun), measurement light intensity (low and
high), \(\log \mathrm{LMA}\), and their interactions on amphistomy
advantage (\(\mathrm{AA}\)). The distributional effects refer to the
effect of factors on residual variation in \(\mathrm{AA}\), denoted
\(\sigma\), on a log-link scale. LMA can also be affected by growth
light intensity and population. For phylogenetically structured random
effects, we report the estimated standard deviation (SD).}

@@

\includegraphics{ms_files/figure-pdf/fig-curves-1.pdf}

>>>>

\bibitem[\citeproctext]{ref-pease_phylogenomics_2016}
\CSLLeftMargin{54. }%
\CSLRightInline{J. B. Pease, D. C. Haak, M. W. Hahn, L. C. Moyle,
\href{https://doi.org/10.1371/journal.pbio.1002379}{Phylogenomics
reveals three sources of adaptive variation during a rapid radiation}.
\emph{PLOS Biology} \textbf{14}, e1002379 (2016).}

\bibitem[\citeproctext]{ref-schoch_dependence_1980}
\CSLLeftMargin{55. }%
\CSLRightInline{P.-G. Schoch, C. Zinsou, M. Sibi,
\href{https://doi.org/10.1093/jxb/31.5.1211}{Dependence of the stomatal
index on environmental factors during stomatal differentiation in leaves
of \emph{{Vigna} sinensis} {L}.: 1. {Effect} of light intensity}.
\emph{Journal of Experimental Botany} \textbf{31}, 1211--1216 (1980).}

\bibitem[\citeproctext]{ref-sack_developmental_2016}
\CSLLeftMargin{56. }%
\CSLRightInline{L. Sack, T. N. Buckley,
\href{https://doi.org/10.1104/pp.16.00476}{The developmental basis of
stomatal density and flux}. \emph{Plant Physiology} \textbf{171},
2358--2363 (2016).}

\bibitem[\citeproctext]{ref-mott_amphistomy_1991}
\CSLLeftMargin{57. }%
\CSLRightInline{K. A. Mott, O. Michaelson, Amphistomy as an adaptation
to high light intensity in \emph{{Ambrosia} cordifolia} ({Compositae}).
\emph{American Journal of Botany} \textbf{78}, 76--79 (1991).}

\bibitem[\citeproctext]{ref-schindelin_fiji_2012}
\CSLLeftMargin{58. }%
\CSLRightInline{J. Schindelin, I. Arganda-Carreras, E. Frise, V. Kaynig,
M. Longair, T. Pietzsch, S. Preibisch, C. Rueden, S. Saalfeld, B.
Schmid, J.-Y. Tinevez, D. J. White, V. Hartenstein, K. Eliceiri, P.
Tomancak, A. Cardona, \href{https://doi.org/10.1038/nmeth.2019}{Fiji: An
open-source platform for biological-image analysis}. \emph{Nature
Methods} \textbf{9}, 676--682 (2012).}

\bibitem[\citeproctext]{ref-monteith_principles_2013}
\CSLLeftMargin{59. }%
\CSLRightInline{J. L. Monteith, M. H. Unsworth, \emph{Principles of
Environmental Physics: Plants, Animals, and the Atmosphere}
(Elsevier/Academic Press, Amsterdam ; Boston, 4th ed., 2013).}

\bibitem[\citeproctext]{ref-stan_development_team_stan_2025}
\CSLLeftMargin{60. }%
\CSLRightInline{Stan Development Team, \emph{Stan {Modeling} {Language}
{Users} {Guide} and {Reference} {Manual}} (2025;
\url{https://mc-stan.org}).}

\bibitem[\citeproctext]{ref-burkner_brms_2017}
\CSLLeftMargin{61. }%
\CSLRightInline{P.-C. Bürkner,
\href{https://doi.org/10.18637/jss.v080.i01}{\textbf{Brms} : {An}
\emph{r} {Package} for {Bayesian} {Multilevel} {Models} {Using}
\emph{stan}}. \emph{Journal of Statistical Software} \textbf{80}
(2017).}

\bibitem[\citeproctext]{ref-gabry_cmdstanr_2025}
\CSLLeftMargin{62. }%
\CSLRightInline{J. Gabry, R. Češnovar, A. Johnson, S. Bronder,
\emph{Cmdstanr: {R} {Interface} to '{CmdStan}'} (2025;
\href{https://mc-stan.org/cmdstanr,\%20https://discourse.mc-stan.org}{https://mc-stan.org/cmdstanr,
https://discourse.mc-stan.org}).}

\bibitem[\citeproctext]{ref-r_core_team_r:_2025}
\CSLLeftMargin{63. }%
\CSLRightInline{R Core Team, \emph{R: {A} {Language} and {Environment}
for {Statistical} {Computing}} (R Foundation for Statistical Computing,
Vienna, Austria, 2025; \url{http://www.R-project.org/}).}

\bibitem[\citeproctext]{ref-gelman_inference_1992}
\CSLLeftMargin{64. }%
\CSLRightInline{A. Gelman, D. B. Rubin, Inference from iterative
simulation using multiple sequences. \emph{Statistical Science}
\textbf{7}, 457--472 (1992).}

@@

We compared \aax{} among 29 ecologically diverse populations of wild
tomato, including representatives of all described species of
\emph{Solanum} sect. \emph{Lycopersicon} and sect.
\emph{Lycopersicoides} (\emph{50}) and the cultivated tomato \emph{S.
lycopersicum} var. \emph{lycopersicum} (Table~\ref{tbl-populations}).
These populations were selected in a previous study because they
encompass the breadth of climatic variation in the wild tomato clade
(\emph{54}). In one case, we substituted a congeric population of the
species (\emph{S. galapagense}) because of difficult growing the focal
population. We also added an a population of \emph{S. pennellii}. Due to
constraints on growth space and time, we spread out measurements over
61.1 weeks from August 29, 2022 to October 31, 2023. Replicates within
population were evenly spread out over this period to prevent
confounding of temporal variation in growth conditions with variation
among populations.

@@

We recorded \(\mathrm{PPFD}\) using full spectrum quantum sensors
(SQ-500-SS, Apogee Instruments, Logan, Utah, USA); we recorded
temperature, RH, and {[}CO\(_2\){]} using an EE894 sensor (E+E
Elektronik, Engerwitzdorf, Austria) protected by a radiation shield. All
environmental measurements were taken every 10 minutes from the middle
of plant racks at approximately the same height as the leaves we
measured. We measured leaf temperature of focal leaves prior to
measurement using an infrared radiometer (SI-111-SS, Apogee Instruments,
Logan, Utah, USA).

@@

Seeds provided by the Tomato Genetics Resource Center germinated on
moist paper in plastic boxes after soaking for 30-60 minutes in a 50\%
(volume per volume) solution of household bleach and water, followed by
a thorough rinse. We transferred seedlings to cell-pack flats containing
Pro-Mix BX potting mix (Premier Tech, Rivière-du-Loup, Quebec, Canada)
once cotyledons fully emerged, typically within 1-2 weeks of sowing. We
grew seeds and seedlings for both sun and shade treatments under the
same environmental conditions (12:12 h,
24.3:\(\qty{21.7}{\degreeCelsius}\), 49.6:58.4 RH day:night cycle). LED
light provided
\(\mathrm{PPFD} = \qty{267}{\micro\mol\raiseto{-2}\meter\raiseto{-1}\second}\)
(Fluence RAZRx, Austin, Texas, USA) during the germination and seedling
stages.

@@

We selected a fully expanded, unshaded leaf at least six leaves above
the cotyledons during early vegetative growth. This typically meant that
plants had grown in light treatments for \(\approx\) 4 weeks, ensuring
they had time to sense and respond developmentally to the light
intensity of the treatment rather than the seedling conditions
(\emph{55}). Shade plants grew slower than sun plants, hence leaves at
the same developmental stage were measured on chronologically older
plants in the shade treatment. In some sun plants, we had to use leaves
higher on the stem because short internodes made lower leaves
inaccessible with the gas exchange equipment. We measured terminal
leaflets in 82.6\% of cases, but used the lateral leaflet closest to the
terminal leaflet when it was damaged or difficult to clamp into the gas
exchange chamber. When a leaflet was damaged during gas exchange
measurements, we collected anatomical data from the nearest leaflet on
the same leaf (1.58\% of leaves).

@@

@@

We estimated the stomatal density and size on ab- and adaxial leaf
surfaces from all leaves, using guard cell length as a proxy for
stomatal size since it is proportional to maximum conductance
(\emph{56}). We made surface impressions of leaf lamina from the same
area used for gas exchange measurements using a-silicone impression
material (Zhermack elite HD+, light body, fast set, Rovigo, Italy). We
applied clear nail polish to make positive replicas of the impression.
After nail polish dried, we mounted replicas on a microscope slide using
transparent tape (\emph{57}). We digitized a portion of each leaf
surface replica using a brightfield microscope (Leica DM2000, Wetzlar,
Germany). We counted and measured guard cell length on all stomata using
the FIJI implementation of ImageJ2 version 2.3.0 (\emph{58}), then
divided the count by the visible leaf area
(\(\qty{0.890}{\raiseto{2}\milli\meter}\)) to estimate stomatal density.
For each surface we calculated the anatomical maximum stomatal
conductance (\gmax) at a reference leaf temperature of
\(\qty{25}{\degreeCelsius}\) following (\emph{56}) as:

@@

where \(D_\mathrm{wv}\) is the diffusion coefficient of water vapor in
air and \(v\) is by the kinematic viscosity of dry air. We assumed
\(D_\mathrm{wv} = \qty{2.49e-5}{\meter\squared\per\second}\) and
\(v = \qty{2.24e-2}{\raiseto{3}\meter\per\mol}\) (\emph{59}). For
kidney-shaped guard cells like those in wild tomatoes,
\(c = h = j = 0.5\). The \gmaxratio{} is the \gmax{} of the adaxial
surface divided by the sum total \gmax{} of both surfaces.

@@

Leaf mass per area (LMA) is the dry mass divided by the leaflet area. We
scanned fresh leaflets on a flat bed scanner (Epson V600, Los Alamitos,
California. USA) and measured leaflet area from digital images using the
FIJI implementation of ImageJ2 version 2.3.0 (\emph{58}). We dried
leaves for 72 hours at \(\qty{74}{\degreeCelsius}\) in a food dehydrator
(Cosori CP267-FD, Vesync Co., Anaheim, California, USA) and weighed
using a benchtop analytical balance (Ohaus PR64 Analytical Balance,
Parsippany, New Jersey, USA). In \(10.5\%\) of leaves we measured LMA on
the adjacent leaflet because the focal leaflet was damaged or wilted
while making surface impressions and we could not reliably estimate
area. LMA data are missing from \(2.97\%\) of individuals because the
area or mass was not recorded at all or recorded incorrectly.

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Estimate \agcurve{} curve parameters
\item
  Estimate \aax{} for both measurement light intensities using
  \agcurve{} curve parameters
\item
  Estimate the effects of light intensity, light treatment, and
  population on \aax{} (acclimation and plasticity hypotheses)
\item
  Estimate the effects of native light habitat on population-level
  \aax{} (constitutive hypothesis)
\end{enumerate}

All models were fit using a Bayesian HMC sampling in the probabilistic
programming language \emph{Stan} (\emph{60}) using the \emph{R} package
\textbf{brms} version 2.22.0 (\emph{61}). We used \emph{CmdStan} version
2.36.0 and \textbf{cmdstanr} version 0.9.0 (\emph{62}) to interface with
\emph{R} version 4.5.1 (\emph{63}). We sampled the posterior
distribution from a single chain with a minimum of 1000 iterations after
1000 warmup iterations. If necessary, we refit models with more
iterations to decrease the Gelman-Rubin convergence statistic
(\(\hat{R}\)) (\emph{64}) to less than 1.01 and the bulk effective
sample size (ESS) to greater than 400 for all model parameters, and
there were fewer than 10 divergent transitions.

@@

We fit Bayesian mixed effects models with phylogenetically structured
random effects to test predictions of competing hypotheses about why
amphistomy advantage (\aax) might be greater for leaves in sunny, open
habitats. We fit all models combinations of measurement light intensity,
growth light intensity, and their interaction as both fixed effects and
random effects among populations. In other words, we test whether there
were main effects of light intensity treatments and/or whether
populations varied in their response to light intensity treatments. In
all models, we included population as a phylogenetically structured
random effect. We accounted for uncertainty in \(\widehat{\mathrm{AA}}\)
using the standard deviation of the posterior distribution of each
\(\widehat{\mathrm{AA}}\) estimate (see above). We crossed fixed and
random effect structures with fixed effects of measurement light
intensity and/or growth light intensity on the distributional parameter,
\(\sigma\), the residual variance in \aax{}. We used a robust regression
approach by assuming \(t\)-distributed residuals, which makes estimates
less susceptible to extreme values. In total, we fit 100 models, each
with a different combination of fixed, random, and distributional
effects. We used the RAxML whole-transcriptome concatenated phylogeny
based on 2,745 100-kb genomic windows from (\emph{54}). Two of our
populations were not in this tree. We used accession LA1044 in place of
LA3909, two populations of \emph{S. galapagense}; LA0750 was added as
sister to LA0716, two closely related populations of \emph{S.
pennellii}. The node separating LA0750 from LA0716 was placed half-way
between the next deepest node. We used the leave-one-out
cross-validation information criterion (LOOIC) to compare the fit of
models (\autoref{tbl-aa_loo1}) using the \emph{R} package \textbf{loo}
version 2.8.0 (\emph{65}) to calculate LOOIC values. We selected the
model with the lowest LOOIC value (Table~\ref{tbl-aa_loo1}) to generate
posterior predictions for hypothesis testing. Parameter estimates from
the selected model are shown in Table~\ref{tbl-fit_aa1}.

To test whether LMA mediated the relationship between growth light
intensity and \aax{}, we repeated a similar model comparison procedure
but included \(\log \mathrm{LMA}\) as an additional fixed effect in the
model, as well its interaction with measurement and growth light
intensity. The model included a direct effect of growth light intensity
on \(\log \mathrm{LMA}\) and random phylogeneticlly structured effects
of population on \(\log \mathrm{LMA}\) and effects of sun treatment on
\(\log \mathrm{LMA}\). This enabled us to disentangle whether
\(\log \mathrm{LMA}\) effects \aax{} directly or indirectly because both
traits respond to growth light intensity. We integrated over missing LMA
data using the \texttt{mi()} function in \textbf{brms}. Because the
evidence from the first set of models strongly supported including
effects of growth and measurement light intensity on the residual
variance in \(\mathrm{AA}\), we included these in terms in all models.
In total, we fit 75 models, each with a different combination of fixed
and random effect structures. We selected the model with the lowest
LOOIC value (Table~\ref{tbl-aa_loo2}) to generate posterior predictions
for hypothesis testing. Note that LOOIC was calculated only from
pointwise likelihood values of the submodel estimating effects of light
intensity, light treatments, and population on \aax{}. Parameter
estimates from the selected model are shown in Table~\ref{tbl-fit_aa2}.

@@

The acclimation, plasticity, and constitutive hypotheses make different
predictions about the relationship between \aax{} and light intensity,
light treatment, and native \(\mathrm{PAI}\) among populations. Since
these hypotheses are not mutually exclusive, we describe how we assessed
support for one, two, or all three hypotheses simultaneously in
\autoref{tbl-predictions}. In general, the acclimation hypothesis was
supported if \aax{} was greater at high light intensity than low light
intensity. The plastic hypothesis was supported if \aax{} was greater in
sun leaves than shade leaves. The constitutive hypothesis was supported
if population-level \aax{} increased with native \ppfd. In interactive
models, we only consider positively reinforcing interactions between
high light intensity, sun leaves, and native \ppfd{} because these are
the only interactions which could explain why amphistomatous leaves are
advantageous in high light habitats.

From the model selected using LOOIC, we predicted the posterior
distribution of the expected \aax{} for each population at each
measurement light intensity and growth light intensity. If the 95\%
confidence intervals for \aax{} did not overlap zero, we considered
\aax{} to be significantly different from zero. We also calculated
posterior distribution for the average \aax{} among populations in each
condition. We tested a linear effect of native light habitat on
population-level \aax{} in each treatment condition, integrating over
the entire posterior distribution to calculate a posterior distribution
of the slope. If the 95\% confidence interval for the slope did not
overlap zero, we considered the slope to be significant.

@@

\subsubsection{The water cost of
hypostomy}\label{the-water-cost-of-hypostomy}

Amphistomy advantage implies that hypostomatous leaves would need to
lose more water by having higher \gsw{} to achieve the same
photosynthetic rate as an otherwise identical amphistomatous leaf. We
estimated this cost of hypostomy approximately. For simplicity, we focus
on a sufficiently small region of the \agcurve{} curve such that we can
approximate the relationship between \gsw{} and \(A\) as log-log linear.
We also assume that the slope is the same in both amphistomatous and
hypostomatous leaves, which was often approximately the case in our
experiment (Figure~\ref{fig-curves}). With these assumptions, the
approximate water cost of hypostomy can be derived from the equations:

@@

\[-\log\left(\frac{g_\text{sw,amphi}}{g_\text{sw,hypo}}\right) = \frac{\mathrm{AA}}{\varepsilon_g}.\]
We note that the slope in this case is the elasticity of \(A\) to \gsw{}
because both are log-transformed. This quantity will be proportional the
water cost of hypostomy assuming that VPD and other environmental
variables are held constant and the boundary layer resistance is
negligible.

@@

\caption{\label{tbl-fit_aa2}Parameter estimates and 95\% confidence
intervals (CIs) from the posterior distribution of the selected model
with leaf mass per area (LMA) as potential mediator variable. Models
potentially include fixed and random effects of growth light intensity
(shade and sun), measurement light intensity (low and high),
\(\log \mathrm{LMA}\), and their interactions on amphistomy advantage
(\(\mathrm{AA}\)). LMA can also be affected by growth light intensity
and population. For phylogenetically structured random effects, we
report the estimated standard deviation (SD).}

@@

\pandocbounded{\includegraphics[keepaspectratio]{ms_files/figure-pdf/fig-curves-1.pdf}}
